manuscript_outline

### 2025-06-21 docs/manuscript/paper.Rmd
deets, dev rate

^2^ Brian's deets here

@@

Development in poikilothermic organisms is strongly temperature-dependent, particularly during early life stages prior to exogenous feeding. This thermally constrained physiology produces a near-mechanistic relationship between ambient temperature and developmental rate, allowing for predictive models of phenology. However, traditional models---often developed under constant laboratory temperatures---struggle to account for the variable thermal regimes experienced by wild populations. To address this, [@sparks2019] introduced the effective value model, a framework that integrates daily temperature data to predict developmental timing under fluctuating conditions. Building on this foundation, the R package **hatchR** was developed to operationalize the effective value model and enable its widespread application, initially focusing on fishes. In this paper, we expand the scope of **hatchR**, demonstrating its applicability across a broad range of poikilothermic taxa, including amphibians, reptiles, and aquatic invertebrates. We also illustrate how this framework supports both applied management goals---such as predicting hatching windows for conservation and monitoring---and basic research on ecological and evolutionary questions, including phenological plasticity, thermal adaptation, and life history strategies. By providing a generalizable, open-source tool, **hatchR** bridges the gap between physiological theory and real-world application, offering a robust platform for predicting developmental phenology in diverse and changing environments.

@@

Because poikilothermic organisms rely on external sources to regulate their body temperature, the physiological processes governing their growth and development are tightly coupled to ambient thermal conditions. This relationship is especially deterministic during early life stages prior to exogenous feeding, when metabolic rates are primarily governed by environmental temperatures, modulated by genetic and maternal effects [@west-eberhard2003]. Although the absolute rate of development can vary widely among species and populations, most poikilotherms exhibit a consistent and often non-linear thermal performance curve that describes development as a function of temperature [@schoolfield1981; @angilletta2006]. These relationships can be effectively approximated with power law or exponential models that capture increasing developmental rates with rising temperature, up to a physiological optimum beyond which performance declines.

Historically, models of developmental phenology in wild populations have relied on the accumulation of degree days (thermal sums) above a species-specific threshold. While widely used, degree-day models are fundamentally linear and fail to account for the non-linear nature of development, leading to biased predictions under fluctuating temperature regimes [@neuheimer2007]. This is particularly problematic for long-developing species or for organisms in variable climates, where embryos incubating at colder temperatures often require fewer thermal units to complete development than those incubating in warmer environments. Moreover, traditional models often assume temporally invariant developmental rates and do not incorporate diel temperature variability, which can significantly affect development in systems with large day--night thermal swings **[CITATION needed, perhaps\*\* <https://doi.org/10.1186/1472-6785-13-18> \*\*...though I wonder why this is brought up, because hatchR doesn't appear to address DTV?, since it auto-sums to daily averages? or am I misunderstanding? It seems like DTV might be better brought up in the discussion as a caveat?]**. Finally, developmental reaction norms may vary across populations due to local adaptation or phenotypic plasticity, further complicating predictions [@sparks2017; @angilletta2009].

To address these limitations in fishes, [@sparks2019] introduced the effective value model---an approach that captures the non-linear, temperature-dependent nature of development by integrating experimental data into a daily developmental rate function. Building on that work, the R package **hatchR** [CITE hatchR] was developed to operationalize the effective value framework and make it accessible to researchers and managers. **hatchR** uses temperature-performance curves generated from lab-based thermal rearing trials to estimate the proportion of development achieved each day under observed temperature conditions. These daily contributions are summed until a cumulative threshold (typically 1.0) is reached, marking the completion of a developmental stage. The package includes both an R-based workflow and a user-friendly Shiny web interface to facilitate broad adoption across research and management communities.

@@

Our results indicate developmental rates were highly variable over different physical habitats (streams originating in lowland hills compared with those originating in mountain ranges with permanent snow fields), with sites in Olympic National Park consistently having the longest development times and the greatest interannual variability (SD \~ 11 days), while lowland streams in the Willapa Hills developed fastest with minimal variability (SD \< 3 days). These results demonstrate the capacity of **hatchR** to detect spatiotemporal phenological variation and forecast responses to shifting climate patterns.

@@

## Phylogenetic Application: Species Differences in Developmental Rates / Reaction Norms

@@

@@

## Local Adaptation: Cogradient Variation in Developmental Rates

@@

Here, we used data from a common garden experiment on cabbage beetles (*Colaphellus bowringi*) reared at six constant temperatures (16--28 °C), spanning populations collected along a \~3,440 km latitudinal gradient in China [@tang2017]. Although all beetles were reared under identical conditions, developmental rates differed among populations, particularly at cooler temperatures. Populations from southern latitudes---where temperatures are generally warmer---developed more quickly than northern populations, a pattern consistent with cogradient variation, where genetic and environmental effects on a phenotype act in the same direction [@conover2009; @sparks2022].

@@

```{r beetles, fig.height=3 , fig.width=5, fig.align='center', fig.cap="Effective value models for five populations of cabbage beetles (Colaphellus bowringi, females) along a latitudinal gradient in China [@tang2017]. Models were parameterized from common garden experiments under constant temperatures. The southernmost population (Xiushui County) shows a consistently faster developmental rate than populations from more northerly sites, especially at colder temperatures. This pattern is indicative of cogradient variation, where both genetic and environmental effects reinforce the observed phenotypic trend."}
tang_temps <- c(16, 19, 22, 24, 26, 28)

@@

We also encourage users to carefully consider how models are parameterized. As discussed in greater detail in Sparks et al. (2025), we recommend using experimental data with at least four distinct temperature treatments spanning a wide thermal range. This is critical for capturing the nonlinear shape of developmental rate curves and avoiding overfitting or extrapolation errors. Many studies we encountered---particularly older or gray literature---relied on only two temperature treatments (e.g., [@qualls1998; @kozák2009]), which we do not recommend for model fitting due to the inability to detect curvature.

>>>>

^2^ Smithsonian Tropical Research Institute, Apartado, Balboa, Ancón, Panamá

@@

Development in poikilothermic organisms is strongly temperature-dependent, particularly during early life stages prior to exogenous feeding. This thermally constrained physiology produces a near-mechanistic relationship between ambient temperature and development rate, allowing for predictive models of phenology. However, traditional models---often developed under constant laboratory temperatures---struggle to account for the variable thermal regimes experienced by wild populations. To address this, [@sparks2019] introduced the effective value model, a framework that integrates daily temperature data to predict developmental timing under fluctuating conditions. Building on this foundation, the R package **hatchR** was developed to operationalize the effective value model and enable its widespread application, initially focusing on fishes. In this paper, we expand the scope of **hatchR**, demonstrating its applicability across a broad range of poikilothermic taxa, including amphibians, reptiles, and aquatic invertebrates. We also illustrate how this framework supports both applied management goals---such as predicting hatching windows for conservation and monitoring---and basic research on ecological and evolutionary questions, including phenological plasticity, thermal adaptation, and life history strategies. By providing a generalizable, open-source tool, **hatchR** bridges the gap between physiological theory and real-world application, offering a robust platform for predicting developmental phenology in diverse and changing environments.

@@

Because poikilothermic organisms rely on external sources to regulate their body temperature, the physiological processes governing their growth and development are tightly coupled to ambient thermal conditions. This relationship is especially deterministic during early life stages prior to exogenous feeding, when metabolic rates are primarily governed by environmental temperatures, modulated by genetic and maternal effects [@west-eberhard2003]. Although the absolute rate of development can vary widely among species and populations, most poikilotherms exhibit a consistent and often non-linear thermal performance curve that describes development as a function of temperature [@schoolfield1981; @angilletta2006]. These relationships can be effectively approximated with power law or exponential models that capture increasing development rates with rising temperature, up to a physiological optimum beyond which performance declines.

Historically, models of developmental phenology in wild populations have relied on the accumulation of degree days (thermal sums) above a species-specific threshold. While widely used, degree-day models are fundamentally linear and fail to account for the non-linear nature of development, leading to biased predictions under fluctuating temperature regimes [@neuheimer2007]. This is particularly problematic for long-developing species or for organisms in variable climates, where embryos incubating at colder temperatures often require fewer thermal units to complete development than those incubating in warmer environments. Moreover, traditional models often assume temporally invariant development rates and do not incorporate diel temperature variability, which can significantly affect development in systems with large day--night thermal swings **[CITATION needed, perhaps\*\* <https://doi.org/10.1186/1472-6785-13-18> \*\*...though I wonder why this is brought up, because hatchR doesn't appear to address DTV?, since it auto-sums to daily averages? or am I misunderstanding? It seems like DTV might be better brought up in the discussion as a caveat?]**. Finally, developmental reaction norms may vary across populations due to local adaptation or phenotypic plasticity, further complicating predictions [@sparks2017; @angilletta2009].

To address these limitations in fishes, [@sparks2019] introduced the effective value model---an approach that captures the non-linear, temperature-dependent nature of development by integrating experimental data into a daily development rate function. Building on that work, the R package **hatchR** [CITE hatchR] was developed to operationalize the effective value framework and make it accessible to researchers and managers. **hatchR** uses temperature-performance curves generated from lab-based thermal rearing trials to estimate the proportion of development achieved each day under observed temperature conditions. These daily contributions are summed until a cumulative threshold (typically 1.0) is reached, marking the completion of a developmental stage. The package includes both an R-based workflow and a user-friendly Shiny web interface to facilitate broad adoption across research and management communities.

@@

Our results indicate development rates were highly variable over different physical habitats (streams originating in lowland hills compared with those originating in mountain ranges with permanent snow fields), with sites in Olympic National Park consistently having the longest development times and the greatest interannual variability (SD \~ 11 days), while lowland streams in the Willapa Hills developed fastest with minimal variability (SD \< 3 days). These results demonstrate the capacity of **hatchR** to detect spatiotemporal phenological variation and forecast responses to shifting climate patterns.

@@

## Phylogenetic Application: Species Differences in Development Rates / Reaction Norms

@@

**BCL NOTE: Maybe I'm misunderstanding, but it doesn't appear to be a strong rank order reversal? From cold to hot, it kinda equalizes, but doesn't clearly reverse?**

@@

## Local Adaptation: Cogradient Variation in Development Rates

@@

Here, we used data from a common garden experiment on cabbage beetles (*Colaphellus bowringi*) reared at six constant temperatures (16--28 °C), spanning populations collected along a \~3,440 km latitudinal gradient in China [@tang2017]. Although all beetles were reared under identical conditions, development rates differed among populations, particularly at cooler temperatures. Populations from southern latitudes---where temperatures are generally warmer---developed more quickly than northern populations, a pattern consistent with cogradient variation, where genetic and environmental effects on a phenotype act in the same direction [@conover2009; @sparks2022].

@@

```{r beetles, fig.height=3 , fig.width=5, fig.align='center', fig.cap="Effective value models for five populations of cabbage beetles (Colaphellus bowringi, females) along a latitudinal gradient in China [@tang2017]. Models were parameterized from common garden experiments under constant temperatures. The southernmost population (Xiushui County) shows a consistently faster development rate than populations from more northerly sites, especially at colder temperatures. This pattern is indicative of cogradient variation, where both genetic and environmental effects reinforce the observed phenotypic trend."}
tang_temps <- c(16, 19, 22, 24, 26, 28)

@@

We also encourage users to carefully consider how models are parameterized. As discussed in greater detail in Sparks et al. (2025), we recommend using experimental data with at least four distinct temperature treatments spanning a wide thermal range. This is critical for capturing the nonlinear shape of development rate curves and avoiding overfitting or extrapolation errors. Many studies we encountered---particularly older or gray literature---relied on only two temperature treatments (e.g., [@qualls1998; @kozák2009]), which we do not recommend for model fitting due to the inability to detect curvature.
